# Application

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
```

The principles outlined in Table 1 are exemplified through the following nursing-focused genomic research example.

## Scenario

Imagine a research team conducting a study to examine whether DNA
methylation (DNAm — a chemical modification to DNA that can influence
gene expression without changing the underlying sequence) in the *ABC*
gene is associated with patient-reported headache outcomes following
aneurysmal subarachnoid hemorrhage (aSAH), a type of stroke caused by
bleeding into the space surrounding the brain.

The study builds on an existing aSAH cohort of 650 patients, with
genome-wide DNAm data available for a subset of 200. The outcomes of
interest are two patient-reported headache measures:

- **Acute headache** — moderate-to-severe headache reported during
  hospitalization
- **Persistent headache** — moderate-to-severe headache reported at
  follow-up

Acute and persistent headache were each coded as binary (0 = absent,
1 = present), with "present" defined as a patient-reported pain score
of ≥4 on an 11-point numeric rating scale (0 = no pain, 10 = worst pain
imaginable), consistent with a moderate-to-severe threshold.
The analytical models examine whether DNAm levels at 13 specific sites
in the *ABC* gene (measured as M-values, a continuous measure of
methylation intensity) predict acute and persistent headache status,
adjusting for age, sex, and Fisher grade (a clinical measure of stroke
severity based on the amount of blood seen on initial brain imaging).

DNAm here reflects peripheral blood methylation, a common and
accessible proxy in biobehavioral nursing research, though not a
direct measure of central pain pathways. In this example study, the *ABC* gene
and the 13 CpG sites were selected a priori based on prior literature and their
location within known regulatory regions. This reflects a candidate-gene approach rather
than exploratory design.

## The Two-Phase Workflow

The proposed workflow spans two phases. Phase one covers planning and
pre-specification; phase two covers code development on synthetic
data, extension to real data, interpretation, and final human review
and disclosure.

![**Figure 3.** Two-phase workflow for responsible human-AI collaboration
in quantitative data analysis.](figures/fig3_two_phase_workflow.png){width=90%}

### Phase One: Planning and Pre-Specification

The first phase establishes the analytical foundation before any code
is written. The researcher articulates the research question, data
structure, planned analytical approach, and known challenges
independently. GenAI is then engaged as a reasoning coach, brought in
to challenge and refine the researcher's thinking, not to generate it.

In this first example, we assign GenAI the role of critical reviewer during planning.

::: {.callout-note title="Our Prompt"}
**Role:** you are a senior biostatistician with expertise in statistical genomics, R programming, and epigenetics.

**Task:** act as a critical reviewer of the analytical plan described below. Identify weaknesses, potential confounders, and power concerns. List the pros and cons of alternative approaches rather than recommending a single fix. Do not write any code.

**Study context:** I am designing a study examining whether DNA methylation (DNAm) at 13 CpG sites in the ABC gene predicts patient-reported headache outcomes following aneurysmal subarachnoid hemorrhage (aSAH). I have mapped out an analytical plan, and the purpose of this interaction is to stress test, refine, and improve it before pre-specifying it with my team. 

**Analytical plan:**

- Outcomes: acute headache and persistent headache...
- Primary predictor: DNAm M-value...
- Covariates: age (continuous)...
- Model: binary logistic regression...
- Multiple testing: Bonferroni correction applied
- Sample: N=650 total; n=200 with DNAm data available
:::

::: {.callout-tip title="AI Response"}
A few things worth stress-testing: (1)
with n=200 for the DNAm subset, statistical power to detect small
methylation effects after Bonferroni correction across 26 tests may be
limited — consider reporting effect sizes and confidence intervals
alongside p-values, not p-values alone; (2) Fisher grade and age may
be correlated with both DNAm patterns and headache severity — confirm
this isn't masking or inflating the DNAm association; (3) an
alternative to per-site Bonferroni correction would be a false
discovery rate (FDR) approach, which is less conservative and common
in epigenome-wide analyses — worth discussing with your biostatistician
which is more defensible for your design.
:::

This kind of exchange is often iterative. Continuing the example above, a follow-up might read:

::: {.callout-note title="Our Prompt"}
Thank you — helpful points. On the power concern: we unfortunately do not have the ability to increase n at this stage, so will proceed with n=200 and plan to report effect sizes and CIs throughout. On the multiple testing question: I discussed FDR vs. Bonferroni with our biostatistician and we decided to use Bonferroni given the candidate-gene design and small number of tests. Can you help me think through whether there are any additional covariates I should consider beyond age, sex, and Fisher grade?
:::

::: {.callout-tip title="AI Response"}
Given the aSAH context, a few additional covariates worth considering: (1) time from aSAH onset to DNAm sample collection, since methylation patterns can shift acutely after neurological injury; (2) medication use at time of sampling, particularly corticosteroids or antiepileptics, which are known to influence DNA methylation; (3) Glasgow Outcome Scale score or clinical severity at discharge, which may confound both headache outcomes and DNAm patterns. Whether to include these depends on availability in your dataset and your pre-specified model — worth flagging in your analytic proposal even if you decide not to adjust for them.
:::

*(The conversation continues.)*

As shown here, any GenAI-generated suggestions should be critically 
evaluated by the researcher, including checking for hallucinations
(plausible-sounding but incorrect information) and cross-referencing
against trusted external sources such as biostatisticians and
published literature. GenAI supplements, but does not replace,
expert consultation.

Once the analytical plan has been refined, it should be written down
as a concise pre-specified analytic proposal and shared with the full
research team for review and approval before any code runs.

### Phase Two: Execution, Translation, and Disclosure

#### Build the Synthetic Dataset

A foundational principle of this workflow is the synthetic-data-first
approach: the full analytical workflow is built and tested using
privacy-safe *simulated data* before any real participant data are
introduced. At no point is real participant data — including
individual rows, values, or screenshots — shared with GenAI platforms.

In this next example, we ask GenAI to generate a synthetic dataset matching our study's structure.

::: {.callout-note title="Our Prompt"}
**Role:** you are an experienced R programmer and biostatistician specializing in epigenetic data workflows. 

**Task:** create an RMarkdown file with annotated code to a generate synthetic dataset matching the structure described below. The purpose of this synthetic dataset is to serve as a privacy-safe substitute for real participant data during analytical code development. Provide inline comments explaining each step and the reasoning behind key distributional choices. Ask me questions if I have missed something that you need in the prompt below. 

**Dataset structure:**

-	Full sample of N=650, with a subset of n=200 having DNAm data available (indicated by a variable called has_dnam)
-	CpG M-values: generate cg1-cg13 only for has_dnam=TRUE participants, drawn from a uniform distribution ranging from -2 to 2; all others should have NA
-	Outcomes: acute_headache (binary, 25% prevalence); persistent_headache (binary, 30% prevalence)
-	Covariates: age (normal distribution, mean=55, SD=10); sex (Male/Female, with Female accounting for 70% of the sample, consistent with the aSAH literature); fisher_grade (categories 2, 3, 4 with probabilities 0.3, 0.4, 0.3)
-	Additional variables: ID (sequential integer, 1-650); has_dnam (logical)

**Variable names and types:** ID (sequential integer, 1-650), age (numeric), sex (character: "male"/"female"), fisher_grade (integer: 2, 3, or 4), acute_headache (binary: 0/1 where 1=yes), persistent_headache (binary: 0/1 where 1=yes), has_dnam (logical: TRUE/FALSE), cg1-cg13 (numeric).

Apply a fixed seed (set.seed(123)) to promote reproducibility.
:::

Note the level of detail in this prompt: every variable, distribution,
and parameter is explicitly defined. These specification decisions are
best made collaboratively with a biostatistician or research team
before writing the prompt, mirroring the pre-specification process
from Phase One. GenAI executes decisions the team has already made; it
does not make them.

::: {.callout-tip title="AI Response"}
Here's R code that builds the cohort
in two stages — first the full N=650 sample with covariates and
outcomes, then the 200-person DNAm subset with CpG M-values — so the
`has_dnam` flag correctly reflects which rows carry methylation data.
:::

Pasting in the GenAI output into our R markdown we can then easily produce a 
fully synthetic dataset! 

```{r generate-synthetic-data}
set.seed(123)  # ensures reproducibility

N <- 650        # full cohort size
n_dnam <- 200   # subset with DNAm data available

# --- Step 1: Generate baseline cohort (N = 650) ---
df <- data.frame(
  ID = 1:N,
  age = round(rnorm(N, mean = 55, sd = 10), 0),
  sex = sample(c("male", "female"), N, replace = TRUE, prob = c(0.3, 0.7)),
  fisher_grade = sample(c(2, 3, 4), N, replace = TRUE, prob = c(0.3, 0.4, 0.3)),
  acute_headache = rbinom(N, 1, 0.25),
  persistent_headache = rbinom(N, 1, 0.30)
)

# --- Step 2: Flag which participants have DNAm data (n = 200 subset) ---
dnam_ids <- sample(df$ID, n_dnam)
df$has_dnam <- df$ID %in% dnam_ids

# --- Step 3: Generate 13 CpG site M-values, DNAm subset only ---
# Participants without DNAm data get NA for all cg1-cg13 columns
cpg_names <- paste0("cg", 1:13)
for (cpg in cpg_names) {
  df[[cpg]] <- NA_real_
  df[[cpg]][df$has_dnam] <- runif(sum(df$has_dnam), min = -2, max = 2)
}

# Preview the structure
str(df[, c("ID", "age", "sex", "fisher_grade", "acute_headache",
           "persistent_headache", "has_dnam", "cg1", "cg2")])
```

```{r preview-data}
head(df)
```

Note that CpG M-values here were simulated from a uniform distribution 
(-2 to 2) purely to support pipeline development and testing, and 
does not reflect the bimodal distribution typically observed in real 
methylation data (M-values usually cluster near the extremes, 
reflecting hypo- and hyper-methylated CpG sites). This simplification 
was a deliberate choice for this synthetic-data stage; it should be 
revisited and validated against the true DNAm distribution once real 
data are available, before any inferential results are reported.

#### Explore the Synthetic Data

Before modeling, confirm the simulated data matches the intended
structure — this check is done by the human, not delegated to GenAI.

```{r describe-data}
# Confirm cohort sizes
table(df$has_dnam)

# Distribution of covariates
summary(df$age)
table(df$sex)
table(df$fisher_grade)

# Outcome prevalence
mean(df$acute_headache)
mean(df$persistent_headache)
```

#### Fit the Analytic Model

Throughout code development, GenAI can be used for targeted tasks, drafting initial code, debugging error messages, refining and annotating existing code, with specific, bounded requests rather than open-ended rewrites. One of the most powerful applications of this targeted approach is using GenAI to build a reusable function, a task that is technically well-defined, involves no participant data, and produces efficient, reproducible code that can be applied across multiple models.

In this next example, we assign GenAI the role of code developer for the primary regression analysis.

::: {.callout-note title="Our Prompt"}
**Role:** you are an experienced R programmer and biostatistician specializing in epigenetic data workflows.

**Task:** using the synthetic dataset generated in our previous exchange, write annotated R code to run the primary regression analyses described below. Build the code in small, testable chunks and include inline comments explaining each step. Do not modify the dataset or introduce new variables.

**Analysis specifications:**

- Run binary logistic regression models for each of 13 CpG sites (cg1–cg13) × 2 outcomes (acute_headache, persistent_headache) = 26 models total; implement this as a reusable R function rather than a loop, to promote code efficiency and reusability
- Model formula for each: outcome ~ CpG_Mvalue + age + sex + fisher_grade, using the DNAm subset only (has_dnam=TRUE)
- For each model, extract the following for the DNAm term only: odds ratio (exponentiated coefficient), 95% confidence interval (exponentiated), and p-value
- Apply Bonferroni correction across the 13 tests within each outcome (corrected significance threshold = 0.05/13 ≈ 0.0038)
- Organize all results into a single clean summary table with columns: CpG site, outcome, odds ratio, lower 95% CI, upper 95% CI, p-value, Bonferroni-corrected significance (yes/no); write that file as an excel document

**Output:** the final summary table should be print-ready and suitable for inclusion in a manuscript.
:::

::: {.callout-tip title="AI Response"}
Here is a reusable function `run_cpg_model()` that fits one logistic regression model given a CpG site name and outcome name, extracts the DNAm term, and returns a tidy results row. Wrapping the logic in a function keeps the code modular and easy to debug one model at a time before scaling to all 26.
:::

Good coding practice is to write a little, run it, check the output,
then build from there. The code below first fits a single
representative model (cg1 → acute headache) to confirm everything
works as intended, then scales up using a reusable function across all
13 CpG sites and both outcomes.

```{r fit-single-model}
# Restrict to the DNAm subset for all models
df_dnam <- subset(df, has_dnam)

# Step 1: Fit one model first to confirm the code works before scaling up
# Logistic regression: acute headache ~ DNAm M-value (cg1) + covariates
model_acute_cg1 <- glm(
  acute_headache ~ cg1 + age + sex + fisher_grade,
  data = df_dnam,
  family = binomial
)

summary(model_acute_cg1)
```

```{r fit-all-models}
# Step 2: Build a reusable function to run one model and extract results
# This is more efficient than copy-pasting code 26 times
run_cpg_model <- function(cpg, outcome, data) {
  
  # Build the model formula dynamically
  formula <- as.formula(paste(outcome, "~", cpg, "+ age + sex + fisher_grade"))
  
  # Fit logistic regression
  mod <- glm(formula, data = data, family = binomial)
  
  # Extract results for the DNAm term only
  coef_table <- summary(mod)$coefficients
  dnAm_row <- coef_table[cpg, ]
  
  # Return a tidy results row
  data.frame(
    cpg_site = cpg,
    outcome = outcome,
    odds_ratio = exp(dnAm_row["Estimate"]),
    ci_lower = exp(dnAm_row["Estimate"] - 1.96 * dnAm_row["Std. Error"]),
    ci_upper = exp(dnAm_row["Estimate"] + 1.96 * dnAm_row["Std. Error"]),
    p_value = dnAm_row["Pr(>|z|)"]
  )
}

# Step 3: Apply the function across all 13 CpG sites and 2 outcomes
cpg_sites <- paste0("cg", 1:13)
outcomes <- c("acute_headache", "persistent_headache")

# Run all 26 models and combine results into a single table
results <- do.call(rbind, lapply(outcomes, function(outcome) {
  do.call(rbind, lapply(cpg_sites, function(cpg) {
    run_cpg_model(cpg, outcome, df_dnam)
  }))
}))

# Step 4: Apply Bonferroni correction within each outcome (0.05/13 ≈ 0.0038)
results$bonferroni_significant <- results$p_value < (0.05 / 13)

# View results
head(results)
```

*(The full results table, including all 26 models, is generated by applying the function across all CpG sites and outcomes as shown above.)*

GenAI can also help us enhance communication by helping draft publication-quality figures. As a bonus example, we assign GenAI the role of figure developer for the forest plot below.

::: {.callout-note title="Our Prompt"}
**Role:** you are an experienced R programmer and biostatistician specializing in epigenetic data workflows.

**Task:** using the `results` object generated in our previous exchange, write annotated R code to create a forest plot displaying the odds ratios and 95% confidence intervals for the DNAm predictor across all 26 models. The plot should:

- Display one row per model (CpG site × outcome combination)
- Facet by outcome (acute_headache vs. persistent_headache)
- Include a vertical reference line at OR = 1.0
- Use three levels of visual emphasis based on significance: gray open circles for non-significant results (p ≥ 0.05), blue filled circles for nominally significant results (p < 0.05), and purple filled circles for Bonferroni-significant results (p < 0.0038); include a legend
- Use clean, publication-ready formatting with minimal theme
:::

::: {.callout-tip title="AI Response"}
Here is ggplot2 code that facets by outcome and uses three levels of visual emphasis: gray open circles for non-significant results, blue filled circles for nominally significant results, and purple filled circles for Bonferroni-significant results.
:::

```{r forest-plot, fig.width=9, fig.height=6}
library(ggplot2)

results_plot <- results
results_plot$cpg_site <- factor(results_plot$cpg_site,
                                 levels = paste0("cg", 13:1))
results_plot$outcome_label <- ifelse(results_plot$outcome == "acute_headache",
                                      "Acute Headache",
                                      "Persistent Headache")

# Flag nominal significance (p < 0.05) separately from Bonferroni
results_plot$significance <- ifelse(results_plot$bonferroni_significant, 
                                     "Bonferroni significant (p < 0.0038)",
                              ifelse(results_plot$p_value < 0.05,
                                     "Nominally significant (p < 0.05)",
                                     "Not significant (p ≥ 0.05)"))

ggplot(results_plot, aes(x = odds_ratio, y = cpg_site,
                          color = significance,
                          shape = significance)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper), height = 0.25) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  facet_wrap(~ outcome_label) +
  scale_color_manual(
    values = c("Not significant (p ≥ 0.05)" = "gray70",
               "Nominally significant (p < 0.05)" = "#2c5f8a",
               "Bonferroni significant (p < 0.0038)" = "#6B2C5F"),
    name = "") +
  scale_shape_manual(
    values = c("Not significant (p ≥ 0.05)" = 1,
               "Nominally significant (p < 0.05)" = 16,
               "Bonferroni significant (p < 0.0038)" = 16),
    name = "") +
  labs(x = "Odds Ratio (95% CI)",
       y = "CpG Site",
       caption = "Dashed line = OR of 1.0. Blue = nominally significant (p < 0.05); purple = Bonferroni significant (p < 0.0038).\nNote: these are synthetic data — no results are expected to reach significance.") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold", color = "#6B2C5F"),
        axis.title = element_text(color = "#6B2C5F"))
```

::: {.callout-important title="Important"}
**Key takeaway:** Real participant data were never shared with GenAI.
The function was built and validated entirely on synthetic data.
When the real dataset is swapped in, this code runs without
modification because it was written for a data structure, not a
specific file.
:::

#### Extend to Real Data

Once the pipeline has been built and validated on synthetic data, it
can be extended to the real data by simply swapping the synthetic data
import for the real one — no other changes to the code are needed.
Because the synthetic and real datasets share identical structure, the
code runs without modification in most cases. When issues arise, GenAI
can assist with debugging by sharing error messages and relevant code,
but never real data.

*(This step is illustrative only in this tutorial, since no real
dataset is used here.)*

#### Interpret Results

Under the guiding principles in Table 1, GenAI is used to translate statistical output into plain language, but the researcher retains full interpretive authority. GenAI functions as a translator of numbers into words, it should not be used to determine clinical or scientific significance. That judgment belongs to the researcher, who brings domain expertise, knowledge of the study population, and understanding of the broader literature that no GenAI tool can replicate.

In this next example, we assign GenAI the role of statistical translator.

::: {.callout-note title="Our Prompt"}
**Role:** you are a senior biostatistician with expertise in statistical genomics, R programming, and epigenetics.

**Task:** translate the statistical results below into plain language including what the odds ratios, confidence intervals, and p-values mean in the context of this specific model, including the direction and magnitude of the association and what the reference category is.

**Boundary:** do not comment on clinical meaning, significance for patient care, or implications for nursing practice; that interpretation belongs to the research team. Limit your response to what the numbers mean statistically.

[paste model output]                      
                                
```{r, echo=FALSE}
summary(model_acute_cg1)
```          
:::

::: {.callout-tip title="AI Response"}
For cg1, the log-odds coefficient is 0.023, which converts to an odds ratio of approximately 1.02 (95% CI: approximately 0.75–1.40). This indicates that for each one-unit increase in DNAm M-value at cg1, the estimated odds of acute headache increase by about 2%, holding all other variables constant. The p-value is 0.89, which is far above any conventional significance threshold. The reference category for sex is 'female' — the sexmale coefficient (0.64, p=0.066) represents the difference in log-odds for males relative to females, converting to an odds ratio of approximately 1.91. Fisher grade has a positive coefficient (0.48, p=0.035), converting to an odds ratio of approximately 1.61 per one-unit increase in grade. Age shows a small negative coefficient (-0.025, p=0.16), converting to an odds ratio of approximately 0.98 per one-year increase.
:::

::: {.callout-important title="Important"}
**Retain interpretive authority:** this prompt illustrates the
'retain interpretive authority' principle in action.
GenAI translates the numbers into words — the researcher decides what
those words mean for patients, populations, and nursing practice. For
researchers at any career stage, making this boundary explicit in the
prompt itself is better than leaving it implicit.
:::

#### Human Review and Disclosure

The workflow closes with human review and transparent disclosure. A
clean, reproducible report is generated from the scripted workflow,
and all code is reviewed line by line by a human. Consistent with the
transparency principle outlined in Table 1, GenAI involvement should
be disclosed in all research products, including the specific tool
used and the nature of its contributions.

